Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2023-01-05 analysis/paper_draft3.Rmd
attempt at adding sentences on incompatibility

In turn, the inference of bias based in part on historical strike data is relevant to lawyers for deciding whether to argue a _Batson_ violation, proving a prima facie case, and ultimately persuading the trial judge that bias in the particular case is more likely than not (Figure \@ref(fig:batsonworkflowfig)). One challenge for incorporating historical data is possible incompatibility between current and historical strike data for a given attorney that may arise because of missing data or because attorney bias depends on certain trial-level characteristics (e.g., defendant race, charge severity).[may need fleshing out]

@@

```
In this paper, we extend a Bayesian approach to _Batson_ and similar challenges [@Kadane2021; @Kadane2018a; @Kadane2018b] to incorporate historical strike data in a manner that allows us to explicitly incorporate the degree of incompatibility between current and historical trials. In this approach, we use two steps. In step one, we specify a model of the peremptory-strike process in the court of interest that includes a bias parameter to which we assign an initial prior distribution. We use an attorney's strike data from past trials, as generated by that same peremptory-strike process, to estimate a posterior distribution for that bias parameter. We use the power prior [@Chen2000; @Ibrahim2015] to control how much the strike history from past cases affects the posterior distribution of the bias parameter. In step two, we estimate the posterior distribution of the bias parameter from the current case, using the posteri-or distribution developed with the historical information and the power prior.

The use of the power prior in this context allows lawyers to explicitly control over the assumed degree of incompatibility between the current and historical trial data. Specifically, lower values (closer to 0) of the power prior parameter, $\alpha$, should be selected when incompatibility is expected to be high, and higher values (closer to 1) of the power prior parameter selected when incompatibility is low. Accordingly, the lawyer bringing the _Batson_ challenge chooses the value of $\alpha$ and must be able to explain and defend that choice as appropriately cooresponding to the likely degree of incompatibility between the current strike data and historical strike data available in the particular case.

>>>>

In turn, the inference of bias based in part on historical strike data is relevant to lawyers for deciding whether to argue a _Batson_ violation, proving a prima facie case, and ultimately persuading the trial judge that bias in the particular case is more likely than not (Figure \@ref(fig:batsonworkflowfig)). One challenge for incorporating historical data is possible incompatibility between current and historical strike data for a given attorney that may arise because of missing data or because attorney bias depends on certain trial-level characteristics (e.g., defendant race, charge severity). Current statistical approaches to the detection of bias have two extreme approaches to inclusion of historical data regardless of incompatibility between current and historical trials. Firstly,  historical data can be ignored entirely. This approach may lead to underpowered estimates of bias due to the low number of strikes in a single trial (typically 6, 10, or 15 strikes per party), and may lead to inflated estimates of bias. Secondly, historical and current trial data can be combined and analysed together. While this approach increases the sample size, it ignores the possible effects of incompatiblity between current and historical trial.

@@

```
In this paper, we extend a Bayesian approach to _Batson_ and similar challenges [@Kadane2021; @Kadane2018a; @Kadane2018b] to incorporate historical strike data in a manner that allows us to explicitly incorporate the degree of incompatibility between current and historical trials. In this approach, we use two steps. In step one, we specify a model of the peremptory-strike process in the court of interest that includes a bias parameter to which we assign an initial prior distribution. We use an attorney's strike data from past trials, as generated by that same peremptory-strike process, to estimate a posterior distribution for that bias parameter. We use the power prior [@Chen2000; @Ibrahim2015] to control how much the strike history from past cases affects the posterior distribution of the bias parameter. In step two, we estimate the posterior distribution of the bias parameter from the current case, using the posterior distribution developed with the historical information and the power prior.

The use of the power prior in this context allows lawyers explicit control over the assumed degree of incompatibility between the current and historical trial data. Specifically, lower values (closer to 0) of the power prior parameter, $\alpha$, should be selected when incompatibility is expected to be high, and higher values (closer to 1) of the power prior parameter selected when incompatibility is low. Accordingly, the lawyer bringing the _Batson_ challenge chooses the value of $\alpha$ and must be able to explain and defend that choice as appropriately cooresponding to the likely degree of incompatibility between the current strike data and historical strike data available in the particular case.
